Post-Disturbance Tree Community Trajectories in a Neotropical Forest

### 2018-05-31 WholePlotTrajectories.Rmd
Discussion

The experiment is a network of twelve 6.25ha plots that underwent a gradient of three logging, thinning and fuelwood cutting treatments (Table \@ref(tab:Tab1)).
Disturbance treatments were attributed according to a randomized plot design with three replicate blocks of four plots. 
The disturbance corresponds to averages of 10 trees removed per hectare with a diameter at 1.3 m height (DBH) above 50 cm for treatment 1 (T1), 32 trees/ha above 40 cm DBH for treatment 2 (T2) and 40 trees above 40 cm DBH for treatment 3 (T3). 
Treatments T2 and T3 besides included the thinning of trees by poison girdling [@Schmitt1989;@Blanc2009]. 
The disturbance intensity was measured as the percentage of aboveground biomass (%AGB) lost between the first inventory in 1984 and five years after disturbance (ref to be found) measured with the BIOMASS R package [@Biomass2018].

@@

The taxonomic diversity was assessed through Richness and the Hill number translation of Shannon and Simpson indices [@Hill1973].
These three indices belong to the set of HCDT or generalized entropy, respectively corresponding to the 0, 1 and 2 order of diversity (q), which proved well suited for diversity studies [@Patil1982;@Tothmeresz1995]. 
The functional diversity was reported using the Rao index of quadratic entropy which combines species abundance distribution and average pairwise dissimilarity based on all functional and life traits.

@@

## Decoupled recovery of communities taxonomic and functional characteristics

Both communities taxonomic and functional diversity and composition proved resilient, following similar humped-back trajectories with a return towards initial values.
The resilience of communities functional characteristics, the most direct link between biodiversity and ecosystem functioning [@Diaz2005], meant a consistent recovery of ecosystem processes in the long term [@Guariguata2001].
The resilience of communities taxonomic characteristics, in turn, meant the maintenance of their initial differences in composition and structure.
It suggested that communities response to disturbance were somehow constraint to converge towards determined compositions [@Hubbell1999; @Molino2001; @Anderson2007; @Baraloto2012a].

Although both communities taxonomic and functional characteristics proved resilient and followed similar humped-back trajectories, the taxonomic recovery systematically lagged behind the corresponding functional dynamics.
Such delay between functional and taxonomic dynamics has already been observed for grasslands [@Tilman1997; @Mouillot2011] and more recently for tropical forests [@Lohbeck2015; @Guariguata2001].
According to the "vegetation quantity effect" [@Grime1998] functional trajectories rely on the pool of dominant species, which diversity and evenness were enhanced after disturbance, and which rapidly restored the functional diversity and composition.
However, communities evenness remained high so infrequent species still missed to the taxonomic recovery that mechanically lagged behind, all the more so that unrecovered species would be functionally redundant and probably undergo competition processes.

@@

Validating the IDH, communities trajectories confirmed the diversity increase after disturbance through an enhanced growth of otherwise less favored species.
This was however completely clear regarding the functional diversity but much more blurry for communities taxonomic characteristics.

The taxonomic richness was weakly or negatively impacted by intense disturbance, as observed on several post logging surveys [@Cannon1998; @Baraloto2012a], while it substantially increased after low intensity disturbance.
Disturbance also enhanced taxonomic evenness, but its intensity was only weakly and lately correlated to communities eveness (*i.e.* Simpson diversity).
Contrastingly, disturbance intensity consistently predicted the significant increase of communities functional diversity for 30 years.
Disturbance impacts involve specific turnover within communities, either among pre-disturbance survivors or among newly recruited trees.
As the composition of old-growth survivors proved to mirror initial communities [@Herault2018], disturbance would impact trees recruitment through the enhanced growth and survival of previously infrequent species and functional characteristics.
Consistently, disturbance resulted in increasing taxonomic dissimilarity compared to pre-disturbance communities and functional shifts towards resource-acquisitive strategies (sharp increase in the SLA, leaf thickness and bark thickness and decrease in wood density, leaf toughness and maximum height) [@Westoby1998; @Wright2004; @Reich2014].
Disturbance then caused a reorganization of the typical high dominance structure of hyperdiverse mature forests after disturbance, benefiting to pioneers and light demanding species. 
Likely, the changes in abiotic environment and competitive pressure favored pioneers which outcompete others in non limiting resources but are excluded in mature forests by long-lived, more resistant and shade tolerant species.
Therefore, consistently with communities dynamics after disturbance relied on species functional strategy and corresponding ability to fill the environmental niches made available by disturbance.
Recruited species then mixed with pre-disturbance ones, from which they differed, and constituted a community all the more diversified that the disturbance was intense [@Molino2001].

## The functional redundancu and the resilience of communities

Both the lag between taxonomic and functional recovery and the middling consistency of the IDH regarding communities taxonomic diversity entail a central role of functional redundancy.
Functional redundancy, determinant of communities resilience [@Trenbath1999; @Elmqvist2003; @Diaz2005], seemed not to have recovered 30 years after disturbance.

Although disturbance would be mitigated by communities functional redundancy, we expect the redundancy itself to be reduced or re-organized within the functional space.
Globally, disturbed communities did not follow consistent trajectories after disturbance and 30 years after disturbance, irrespective of the intensity, disturbed communities displayed lower, similar or higher functional redundancy.
The redundancy within the functional space of the initial community however clearly followed humped-back trajectories which decrease was determined by the disturbance intensity and which return to initial state was not observed for any disturbed community.
Not only did the functional redundancy and hence the resilience of pre-disturbance community decrease, but also did the ones of disturbance-specific communities increase.
It means higher chances to have long lasting or self-maintained compositional changes towards disturbance resistant species, lianas or epiphytes [@Haddad2008; @Burslem2000; @Martin2013] and thus highly question forest's resilience [@Chazdon2003a].
Specificaly, this would impair species contingent to undisturbed forests, threatening their maintenance, and run the risk to loose cornerstone species and trigger unexpected ecological consequences [@Jones1994;@Diaz2005;@Gardner2007].

@@

our study disentangled tropical forests functional and taxonomic decoupled response to disturbance, with a rapid recovery of communities functioning but slower and more variable taxonomic dynamics.
Consistently with the IDH functional trajectories were constrained by environmental pressures favoring pioneers and light-demanding species after disturbance, but taxonomic dynamics were more stochastic as they involved infrequent speciesfunctoinally redundant with the dominants and thus under competitive pressure.
While communities functioning rapidly recovered, taxonomic trajectories, although consistent and converging towards full recovery, were much slower and entailed persisting alteration of communities functional redundancy.
the trajectories therefore suggested a potential sustainability of tropical forests in the face of quite intense disturbance but only when followed by long recovery periods, longer than 30 years [@Gourlet-Fleury2005].
This study besides highlighted the central role of recruitment processes underlying the IDH mechanisms and holding the whole community response to disturbance.

>>>>

The experiment is a network of twelve 6.25ha plots that underwent a gradient of three logging, thinning and fuelwood cutting treatments (Table \@ref(tab:Tab1)).
Disturbance treatments were attributed according to a randomized plot design with three replicate blocks of four plots. 
The disturbance corresponds to averages of 10 trees removed per hectare with a diameter at 1.3 m height (DBH) above 50 cm for treatment 1 (T1), 32 trees/ha above 40 cm DBH for treatment 2 (T2) and 40 trees/ha above 40 cm DBH for treatment 3 (T3). 
Treatments T2 and T3 besides included the thinning of trees by poison girdling [@Schmitt1989;@Blanc2009]. 
The disturbance intensity was measured as the percentage of aboveground biomass (%AGB) lost between the first inventory in 1984 and five years after disturbance (ref to be found) measured with the BIOMASS R package [@Biomass2018].

@@

The taxonomic diversity was assessed through Richness and the Hill number translation of Shannon and Simpson indices [@Hill1973].
These three indices belong to the set of HCDT or generalized entropy, respectively corresponding to the 0, 1 and 2 order of diversity (q), recomended for diversity studies [@Patil1982;@Tothmeresz1995]. 
The functional diversity was reported using the Rao index of quadratic entropy which combines species abundance distribution and average pairwise dissimilarity based on all functional and life traits.

@@

## Decoupled taxonomic and functional trajectories

Both communities taxonomic and functional diversity and composition proved resilient, following similar humped-back trajectories returning towards initial values.
The resilience of communities functional structure, that is the most direct link between biodiversity and ecosystem functioning [@Diaz2005], meant a consistent recovery of ecosystem processes in the long term [@Guariguata2001].
The resilience of communities taxonomy meant the maintenance of their initial differences in composition and diversity, suggesting the existence of multiple stable equilibria as assumed for highly diverse and productive ecosystems [@Chase2003].
It besides suggested that recovery trajectories depend on the initial composition eventually restored that would the the asymptot of the humped-shaped dynamics [@Hubbell1999; @Molino2001; @Anderson2007; @Baraloto2012a].

Although both communities taxonomic and functional characteristics proved resilient and followed similar humped-back trajectories, the taxonomic recovery systematically lagged behind the functional.
The decoupling between functional and taxonomic dynamics has already been observed for grasslands [@Tilman1997; @Mouillot2011] and more recently for tropical forests [@Lohbeck2015; @Guariguata2001].
According to the "vegetation quantity effect" [@Grime1998] functional trajectories rely on the pool of dominant species, which diversity and evenness were enhanced after disturbance, and which rapidly recovered their initial functional structure.
However, communities evenness remained high so infrequent species still missed to the taxonomic recovery that proved all the more longer to recover that unrecovered species would be functionally redundant and probably underwent competition processes.

@@

Validating the IDH, communities trajectories confirmed the diversity increase after disturbance through an enhanced growth of otherwise less favored species.
This was strong regarding the functional diversity but more blurry regarding their taxonomic structure.

The taxonomic richness was weakly or negatively impacted by intense disturbance, as observed on several post logging surveys [@Cannon1998; @Baraloto2012a], while it substantially increased after low intensity disturbance.
Disturbance also enhanced taxonomic evenness, but the disturbance intensity was only weakly and lately correlated to communities eveness (*i.e.* Simpson diversity).
Contrastingly, disturbance intensity consistently predicted the significant increase of communities functional diversity for 30 years.
Disturbance entailed specific turnover within communities, either among pre-disturbance survivors or among newly recruited trees.
As the composition of old-growth survivors proved to mirror initial communities [@Herault2018], disturbance likely impacted trees recruitment through the enhanced growth and survival of previously infrequent species and functional types.
Consistently, disturbance resulted in increasing taxonomic dissimilarity compared to pre-disturbance communities and significant functional shifts towards resource-acquisitive strategies (sharp increase in the SLA, leaf thickness and bark thickness and decrease in wood density, leaf toughness and maximum height) [@Westoby1998; @Wright2004; @Reich2014].
Disturbance then caused a reorganization of the typical high dominance structure of hyperdiverse mature forests after disturbance, benefiting to pioneers and light demanding species. 
Likely, the changes in abiotic environment and competitive pressure favored pioneers which outcompete other species in non limiting resources but are excluded in mature forests through the cmpetition with long-lived, more resistant and shade bearers.
Therefore, consistently with the IDH the trajectories after disturbance relied on the environmental niches made available by disturbance filled by species from a restrictied functional range.
Recruited species therefore differed from pre-disturbance ones, constituting a community all the more diversified that the disturbance was intense [@Molino2001].

## The functional redundancy, key of communities resilience

Both the lag between taxonomic and functional recovery and the middling consistency of the IDH regarding communities taxonomic diversity made central the role of functional redundancy.
Functional redundancy, determinant of communities resilience [@Trenbath1999; @Elmqvist2003; @Diaz2005], seemed not to have recovered 30 years after disturbance.

Communities with high functional redundancy are expected to be able to lose species with a limited impact on their functional structure [@Carmona2016].
Functional redundancy therefore mitigates the impacts of disturbance, but we expect the redundancy itself to be reduced or re-organized within the functional space.
Although the overall functional redundancy did not follow consistent trajectories after disturbance, the redundancy restricted to the functional space of the initial community clearly followed humped-back trajectories.
This restricted redundancy decreased according to the disturbance intensity before returning towards the initial value but had not recovered for any disturbed community after 30 years.
The recovery of the redundancy within the initial functional space explained the hampered recovery of infrequent species specific to old-growth forests that would be functionally redundant and therefore underwent competitive pressures limiting species functional similarity [@Mayfield2010].
The unachieved humped-shape redundancy trajectory meant a decrease of the resilience of pre-disturbance communities and an increase in the resilience of disturbance-specific communities (corresponding to lower richness and higher dominance of pioneer species).
After disturbance the chances were higher to have long lasting or self-maintained compositional changes in favor of disturbance resistant species, lianas or epiphytes [@Haddad2008; @Burslem2000; @Martin2013], which highly questioned the completeness of the resilience [@Chazdon2003a].
Specificaly, species contingent to undisturbed forests might be impaired, threatening their maintenance, the risk to loose cornerstone species and trigger unexpected ecological consequences would be higher [@Jones1994;@Diaz2005;@Gardner2007].

@@

our study defined the decoupled functional and taxonomic trajectories of tropical rainforests after disturbance.
These trajectories shed light on their rapid functional recovery but slower and more variable taxonomic dynamics.
Consistently with the IDH, functional trajectories were constrained by determined processes favoring pioneers and light-demanding species after disturbance.
The taxonomic dynamics however proved more stochastic due to the recovery of infrequent, functionally redundant species hampered by limiting similarity processes.
Although communities functioning rapidly recovered, their taxonomic structured and functional redundancy remained altered for several decades.
The trajectories therefore suggested a potential sustainability of tropical forests in the face of quite intense disturbance but only when followed by long recovery periods, longer than 30 years [@Gourlet-Fleury2005].
This study besides highlighted the central role of recruitment processes underlying the IDH mechanisms and holding the whole community response to disturbance.
